---
title: "Data preparation for regression analysis_Sensitivity 30 days"
author: "Huiru Dong"
date: "2024-04-30"
output: html_document
---


```{r}
# Load all the libraries
library(data.table)
library(dplyr)

# Folder that store the data
dat.folder = "C:\\Users\\hd820\\OneDrive - Mass General Brigham\\Data Repository\\IQVIA_analysis\\[P7] Buprenorphine dosage\\Data\\"
```


Read and check the episode and RXs records
```{r}
# Read the full dataset: episode merged with RXs records
RX_episode = fread(paste(dat.folder,"episodes_RX_with_max_dosage_M18_Sensitivity Day30.csv", sep=''))
setorder(RX_episode,PATIENT_ID,episode.start,SVC_DT)
# Check number of records
nrow(RX_episode)
# 15,606,863

# Check unique number of episodes
length(unique(RX_episode$uuid))
# 1,213,812 episodes
```


```{r}
# Check how many episodes involved a max dosage variable >40
check=RX_episode[RX_episode$max_dosage_1M>40 | RX_episode$max_dosage_3M>40 | RX_episode$max_dosage_6M>40 |
                RX_episode$max_dosage_12M>40 | RX_episode$max_dosage_18M>40,]
check2 = RX_episode[RX_episode$uuid %in% check$uuid,c("PATIENT_ID","episode.start","episode.end","SVC_DT_date","DAYS_SUPPLY_CNT","dosage","max_dosage_1M","max_dosage_3M","max_dosage_6M","max_dosage_12M","max_dosage_18M")]
length(unique(check$uuid))
# 4242 episodes with max dosage >40 due to overlapping days
nrow(check2)
# 192399 rows
```


SAMPLE RESTRICTION:
Remove these max dosage>40 episodes
```{r}
RX_episode2 = RX_episode[!(RX_episode$uuid %in% check$uuid),]
nrow(RX_episode2)
# 15414464
length(unique(RX_episode2$uuid))
# 1209570
```

SAMPLE RESTRICTION:
New restriction: check how many episode started before April 1, 2020
```{r}
nrow(RX_episode2[RX_episode2$episode.start < "2020-04-01",])
# 14182620 before Apr 1, 2020

RX_episode3 = RX_episode2[RX_episode2$episode.start < "2020-04-01",]

# Double check
nrow(RX_episode3)
min(unique(RX_episode3$episode.start))
max(unique(RX_episode3$episode.start))
length(unique(RX_episode3$uuid))
# 1018638
```
After sample restriction (Apr 1), we have 1018638 episodes 14182620 records



POTENTIAL SAMPLE RESTRICTION:
We will define an indicator: 180 days washout period, only include episodes without treatment within 180 days
```{r}
episode.ID = RX_episode3[,c("PATIENT_ID","episode.start", "episode.end")]
episode.ID = episode.ID[!duplicated(episode.ID)]
setorder(episode.ID, PATIENT_ID,episode.start)

# For patients with multiple episode, look at the gap between episode 
episode.ID = episode.ID %>%
  group_by(PATIENT_ID) %>%
  mutate(previous_end = lag(episode.end, n=1, order_by=PATIENT_ID))

episode.ID$lag = ifelse(is.na(episode.ID$previous_end),99999,
                        (episode.ID$episode.start-episode.ID$previous_end))
episode.ID$keep_washout180 = ifelse(episode.ID$lag>180,1,0)

table(episode.ID$keep_washout180, useNA="always")

# Keep only useful variables
episode.ID = episode.ID[,c("PATIENT_ID","episode.start","keep_washout180")]

```

Keep the analytical episode sample data
```{r}
# keep only one record per episode, no need to have all RXs 
episode = RX_episode3 %>%
   group_by(PATIENT_ID, episode.start) %>%
   arrange(PATIENT_ID,episode.start,SVC_DT) %>%
   filter(row_number()==1)
nrow(episode)

# Merge in the washout indicator
episode = merge(episode, episode.ID, by=c("PATIENT_ID","episode.start"), all.x=TRUE)

# keep useful variables
episode.mod = episode[,c("PATIENT_ID","episode.start","episode.end","episode.duration","PAY_TYP_DESC",
                         "max_dosage_1M","max_dosage_3M","max_dosage_6M","max_dosage_12M","max_dosage_18M",
                         "keep_washout180")]
setorder(episode.mod, PATIENT_ID, episode.start)
```
```{r}
# Clean up
rm(check, check2, episode.ID, episode, RX_episode, RX_episode2)
```

Double check the max dosage variables 
```{r}
# For episodes lasted <1M, max dosage at 1M, 3M, 6M, 12M, and 18M should be equal
check.1M = episode.mod[episode.mod$episode.duration<30,]
check.1M$mean = (check.1M$max_dosage_1M+
                    check.1M$max_dosage_3M+
                    check.1M$max_dosage_6M+
                    check.1M$max_dosage_12M+
                    check.1M$max_dosage_18M)/5
check.1M$flag = ifelse(round(check.1M$mean,2)!=round(check.1M$max_dosage_1M,2), 1,0)
table(check.1M$flag, useNA="always")

# For episodes lasted <3M, max dosage at 3M, 6M, 12M, 18M should be equal
check.3M = episode.mod[episode.mod$episode.duration<30*3,]
check.3M$mean = (check.3M$max_dosage_3M+
                 check.3M$max_dosage_6M+
                 check.3M$max_dosage_12M+
                 check.3M$max_dosage_18M)/4
check.3M$flag = ifelse(round(check.3M$mean,2)!=round(check.3M$max_dosage_3M,2), 1,0)
table(check.3M$flag, useNA="always")

# For episodes lasted <6M, max dosage at 6M, 12M, 18M should be equal
check.6M = episode.mod[episode.mod$episode.duration<30*6,]
check.6M$mean = (check.6M$max_dosage_6M+
                 check.6M$max_dosage_12M+
                 check.6M$max_dosage_18M)/3
check.6M$flag = ifelse(round(check.6M$mean,2)!=round(check.6M$max_dosage_6M,2), 1,0)
table(check.6M$flag, useNA="always")

# For episodes lasted <12M, max dosage at 12M and 18M should be equal
check.12M = episode.mod[episode.mod$episode.duration<30*12,]
check.12M$flag = ifelse(check.12M$max_dosage_12M!=check.12M$max_dosage_18M, 1,0)
table(check.12M$flag, useNA="always")
```


Merge in patient demographic data
```{r}
# Read in patient data
d_patient = fread(file = "\\\\rfawin.partners.org\\mgh-itafisma\\IQVIA\\Patient.gz",
                  select = c("PATIENT_ID","PAT_BRTH_YR_NBR","PAT_GENDER_CD","PAT_ETHNICITY"))
setorder(d_patient, PATIENT_ID)

# Merge patient information with episode data
episode.mod2 = merge (episode.mod, d_patient, by = c("PATIENT_ID"), all.x = TRUE)

```



SAMPLE RESTRICTION
Remove records with missing sex 
```{r}
table(episode.mod2$PAT_GENDER_CD, useNA="always")
episode.mod3 = episode.mod2[episode.mod2$PAT_GENDER_CD!="U",]
table(episode.mod3$PAT_GENDER_CD, useNA="always")
nrow(episode.mod3)
```

SAMPLE RESTRICTION
Restrict the episodes to those with age 18-85
```{r}
episode.mod3$age = year(episode.mod3$episode.start)-episode.mod3$PAT_BRTH_YR_NBR
summary(episode.mod3$age)
nrow(episode.mod3[episode.mod3$age<18 | is.na(episode.mod3$age),])

# Clean up the age variable, keeping 18-85 years old
episode.mod4 = episode.mod3[episode.mod3$age>=18 & episode.mod3$age<=85 & !is.na(episode.mod3$age),]
nrow(episode.mod4)

# Create age categories 
episode.mod4$age_cat = 
  ifelse(episode.mod4$age>=18 & episode.mod4$age<=24,"1. 18-24",
    ifelse(episode.mod4$age>=25 & episode.mod4$age<=34,"2.25-34",
           ifelse(episode.mod4$age>=35 & episode.mod4$age<=44, "3.35-44",
                  ifelse(episode.mod4$age>=45 & episode.mod4$age<=64,"4.45-64",
                         "5.65+"))))
table(episode.mod4$age_cat, useNA="always")
summary(episode.mod4$age)
```

Clean up the race/ethnicity variable
```{r}
table(episode.mod4$PAT_ETHNICITY, useNA="always")
# Combine unknown and unspecified
episode.mod4$race = ifelse(episode.mod4$PAT_ETHNICITY %in% c("UNKNOWN","UNSPECIFIED"),"UNKNOWN",episode.mod4$PAT_ETHNICITY)

table(episode.mod4$race, episode.mod4$PAT_ETHNICITY,useNA="always")
```

Check payment type
```{r}
table(episode.mod4$PAY_TYP_DESC, useNA="always")
episode.mod4$payment = ifelse(episode.mod4$PAY_TYP_DESC %in% c("MEDICARE PART D","MEDICARE"),
                              "MEDICARE",episode.mod4$PAY_TYP_DESC)
table(episode.mod4$payment,episode.mod4$PAY_TYP_DESC, useNA="always")
```

Generate year variable
```{r}
episode.mod4$year = year(episode.mod4$episode.start)
```

Generate outcome variable
```{r}
episode.mod4$outcome.1M = ifelse(episode.mod4$episode.duration>=30,1,0)
episode.mod4$outcome.3M = ifelse(episode.mod4$episode.duration>=30*3,1,0)
episode.mod4$outcome.6M = ifelse(episode.mod4$episode.duration>=30*6,1,0)
episode.mod4$outcome.12M = ifelse(episode.mod4$episode.duration>=30*12,1,0)
episode.mod4$outcome.18M = ifelse(episode.mod4$episode.duration>=30*18,1,0)


# Double check
table(episode.mod4$outcome.1M, useNA="always")
table(episode.mod4$outcome.3M, useNA="always")
table(episode.mod4$outcome.6M, useNA="always")
table(episode.mod4$outcome.12M, useNA="always")
table(episode.mod4$outcome.18M, useNA="always")


summary(episode.mod4[episode.mod4$outcome.1M==1,]$episode.duration)
summary(episode.mod4[episode.mod4$outcome.3M==1,]$episode.duration)
summary(episode.mod4[episode.mod4$outcome.6M==1,]$episode.duration)
summary(episode.mod4[episode.mod4$outcome.12M==1,]$episode.duration)
summary(episode.mod4[episode.mod4$outcome.18M==1,]$episode.duration)
```


Generate max dosage categorical variables
```{r}

# Create categorical variable for the 1M, 3M, 6M, 12M, and 18M variables
episode.mod4$max_dosage_1M_cat = 
  cut(episode.mod4$max_dosage_1M, 
      breaks = c(0, 4, 8, 12, 16, 20, 24, 28, 32, 36, 40),
      labels = c('1.0-4', '2.5-8', '3.9-12', '4.13-16', '5.17-20', '6.21-24', '7.25-28', '8.29-32', '9.33-36', '10.37-40'))

# Double check
summary(episode.mod4$max_dosage_1M)
table(episode.mod4$max_dosage_1M_cat, useNA="always")
summary(episode.mod4[episode.mod4$max_dosage_1M_cat=="4.13-16",]$max_dosage_1M)


episode.mod4$max_dosage_3M_cat = 
  cut(episode.mod4$max_dosage_3M, 
      breaks = c(0, 4, 8, 12, 16, 20, 24, 28, 32, 36, 40),
      labels = c('1.0-4', '2.5-8', '3.9-12', '4.13-16', '5.17-20', '6.21-24', '7.25-28', '8.29-32', '9.33-36', '10.37-40'))

# Double check
summary(episode.mod4$max_dosage_3M)
table(episode.mod4$max_dosage_3M_cat, useNA="always")
summary(episode.mod4[episode.mod4$max_dosage_3M_cat=="4.13-16",]$max_dosage_3M)


episode.mod4$max_dosage_6M_cat = 
  cut(episode.mod4$max_dosage_6M, 
      breaks = c(0, 4, 8, 12, 16, 20, 24, 28, 32, 36, 40),
      labels = c('1.0-4', '2.5-8', '3.9-12', '4.13-16', '5.17-20', '6.21-24', '7.25-28', '8.29-32', '9.33-36', '10.37-40'))

# Double check
summary(episode.mod4$max_dosage_6M)
table(episode.mod4$max_dosage_6M_cat, useNA="always")
summary(episode.mod4[episode.mod4$max_dosage_6M_cat=="4.13-16",]$max_dosage_6M)


episode.mod4$max_dosage_12M_cat = 
  cut(episode.mod4$max_dosage_12M, 
      breaks = c(0, 4, 8, 12, 16, 20, 24, 28, 32, 36, 40),
      labels = c('1.0-4', '2.5-8', '3.9-12', '4.13-16', '5.17-20', '6.21-24', '7.25-28', '8.29-32', '9.33-36', '10.37-40'))

# Double check
summary(episode.mod4$max_dosage_12M)
table(episode.mod4$max_dosage_12M_cat, useNA="always")
summary(episode.mod4[episode.mod4$max_dosage_12M_cat=="4.13-16",]$max_dosage_12M)


episode.mod4$max_dosage_18M_cat = 
  cut(episode.mod4$max_dosage_18M, 
      breaks = c(0, 4, 8, 12, 16, 20, 24, 28, 32, 36, 40),
      labels = c('1.0-4', '2.5-8', '3.9-12', '4.13-16', '5.17-20', '6.21-24', '7.25-28', '8.29-32', '9.33-36', '10.37-40'))

# Double check
summary(episode.mod4$max_dosage_18M)
table(episode.mod4$max_dosage_18M_cat, useNA="always")
summary(episode.mod4[episode.mod4$max_dosage_18M_cat=="4.13-16",]$max_dosage_18M)

```

```{r}
# Clean up
rm(d_patient, episode.mod, episode.mod2,episode.mod3,
   check.1M,check.3M,check.6M,check.12M)
```


```{r}
# Save the data file for regression model
fwrite(episode.mod4, paste(dat.folder,"Model data_18M_Sensitivity Day30.csv", sep=''))
# 1,011,968

```





