---
title: "No change dosage"
author: "Huiru Dong"
date: "2024-04-26"
output: html_document
---

```{r}
# Load all the libraries
library(data.table)
library(dplyr)
library(emmeans)
library(tibble)

# Folder that store the data
dat.folder = "C:\\Users\\hd820\\OneDrive - Mass General Brigham\\Data Repository\\IQVIA_analysis\\[P7] Buprenorphine dosage\\Data\\"
result.folder = "C:\\Users\\hd820\\OneDrive - Mass General Brigham\\Data Repository\\IQVIA_analysis\\[P7] Buprenorphine dosage\\Result\\"
```


```{r}
# Read in the data
dat = fread(paste(dat.folder,"Model data_18M.csv", sep=''))
nrow(dat)
```


SAMPLE RESTRICTION:
Keep the 180 washout period
```{r}
# Check the frequencies for the washout indicator 
table(dat$keep_washout180, useNA="always")

# Keep the washout indicator=1
dat = dat[dat$keep_washout180==1,]
nrow(dat)
# 704446
```


ADDITIONAL RESTRICTION
```{r}
# Only keep those didn't change dosage throughout the investigated period 
nochange = dat %>%
  filter(max_dosage_1M_cat == max_dosage_3M_cat & 
         max_dosage_3M_cat == max_dosage_6M_cat & 
         max_dosage_6M_cat == max_dosage_12M_cat & 
         max_dosage_12M_cat == max_dosage_18M_cat)
# 620229

nrow(nochange)/nrow(dat)*100
# 88% did not change

# Number of unique participants
length(unique(nochange$PATIENT_ID))
```

Check and regroup the dose variables
```{r}
# table(nochange$max_dosage_1M_cat, useNA="ifany")
# table(nochange$max_dosage_3M_cat, useNA="ifany")
# table(nochange$max_dosage_6M_cat, useNA="ifany")
# table(nochange$max_dosage_12M_cat, useNA="ifany")
# table(nochange$max_dosage_18M_cat, useNA="ifany")

# Regrouping for a single variable
nochange <- nochange %>%
  mutate(
    max_dosage_1M_cat_grouped = case_when(
      max_dosage_1M_cat %in% c("7.25-28", "8.29-32") ~ "7.25-32",
      max_dosage_1M_cat %in% c("9.33-36", "10.37-40") ~ "8.33-40",
      TRUE ~ max_dosage_1M_cat # Retain other categories
    )
  )

# List of variables to regroup
variables <- c('max_dosage_1M_cat', 'max_dosage_3M_cat', 
               'max_dosage_6M_cat', 'max_dosage_12M_cat', 
               'max_dosage_18M_cat')

# Apply regrouping to each variable
nochange <- nochange %>%
  mutate(across(all_of(variables), ~ case_when(
    . %in% c("7.25-28", "8.29-32") ~ "7.25-32",
    . %in% c("9.33-36", "10.37-40") ~ "8.33-40",
    TRUE ~ . # Retain other categories
  ), .names = "{.col}_grouped"))

# Double check
table(nochange$max_dosage_1M_cat_grouped, nochange$max_dosage_1M_cat)
table(nochange$max_dosage_3M_cat_grouped, nochange$max_dosage_3M_cat)
table(nochange$max_dosage_6M_cat_grouped, nochange$max_dosage_6M_cat)
table(nochange$max_dosage_12M_cat_grouped, nochange$max_dosage_12M_cat)
table(nochange$max_dosage_18M_cat_grouped, nochange$max_dosage_18M_cat)

```


Change reference levels for key variables
```{r}
# Dosage categories
# Define the common levels
common_levels = c('4.13-16', '1.0-4', '2.5-8', '3.9-12', '5.17-20', '6.21-24', '7.25-32', '8.33-40')

# List of column names to transform
columns_to_transform = c('max_dosage_1M_cat_grouped', 'max_dosage_3M_cat_grouped', 
                         'max_dosage_6M_cat_grouped', 'max_dosage_12M_cat_grouped', 
                         'max_dosage_18M_cat_grouped')

# Apply factor transformation with common levels to each specified column
for (col in columns_to_transform) {
  nochange[[col]] = factor(nochange[[col]], levels = common_levels)
}


# Models adjust for race, PAT_GENDER_CD, payment, year
nochange$race = factor(nochange$race,
                  levels=c("WHITE","BLACK","HISPANIC","ASIAN - OTHER","UNKNOWN"))

nochange$PAT_GENDER_CD = factor(nochange$PAT_GENDER_CD, levels=c("M","F"))

nochange$payment = factor(nochange$payment, levels=c("THIRD PARTY","CASH","MEDICAID","MEDICARE"))

# double check
# table(nochange$max_dosage_18M_cat, useNA="always")

```


Remove datasets that are no longer needed
```{r}
rm(dat, col, columns_to_transform, common_levels, variables)
```




Bivariable logistic regression
```{r}
# Define the outcome periods and corresponding dosage categories
outcome_periods <- c("1M", "3M", "6M", "12M", "18M")
dosage_categories <- c("max_dosage_1M_cat_grouped", "max_dosage_3M_cat_grouped", 
                       "max_dosage_6M_cat_grouped", "max_dosage_12M_cat_grouped",
                       "max_dosage_18M_cat_grouped")

# Define the predictor variables (excluding dosage category)
common_predictors <- c("age_cat", "race", "PAT_GENDER_CD", "payment", "year")

# Initialize an empty list to store all results
all_results <- list()

# Loop over each outcome period
for (i in seq_along(outcome_periods)) {
  outcome <- outcome_periods[i]
  dosage <- dosage_categories[i]
  
  # Define the predictor variables including the specific dosage category
  predictor_vars <- c(dosage, common_predictors)

  # Iterate over each predictor variable
  for (var in predictor_vars) {
    # Fit bivariable logistic regression model
    formula <- as.formula(paste0("outcome.", outcome, " ~ ", var))
    model <- glm(formula, data = nochange, family = "binomial")
    
    # Extract coefficients and confidence intervals
    coef_raw <- summary(model)$coefficients
    coef_vals <- exp(coef(model))
    conf_int <- exp(confint.default(model))
    
    # Combine coefficients and confidence intervals into a data frame
    result_df <- data.frame(
      "Predictor" = var,
      "Outcome_Period" = outcome,
      "Odds ratio" = coef_vals,
      "CI_lower" = conf_int[,1],
      "CI_upper" = conf_int[,2],
      "Coefficient" = coef_raw[,1],
      "P" = coef_raw[,4]
    )
    
    # Store the results in the list
    all_results <- append(all_results, list(result_df))
    
    # Clean up intermediate variables to save memory
    rm(model, formula, coef_raw, coef_vals, conf_int)
  }
}

# Combine all results into a single data frame
combined_results <- do.call(rbind, all_results)

# Save the combined results as a CSV file
combined_output_filename <- paste0(result.folder, "combined_bivariable_M18_nochangedosage_revision.csv")
write.csv(combined_results, combined_output_filename, row.names = TRUE)

# Clean up memory
rm(all_results, combined_results, common_predictors, outcome_periods, dosage_categories, predictor_vars,
   var,outcome,i,dosage,combined_output_filename,result_df)

```


```{r}
# Double check
mod.test = glm(outcome.18M~max_dosage_18M_cat_grouped, data=nochange, family="binomial")
summary(mod.test)
rm(mod.test)
```






Multivariable logistic regression
```{r}
# Define the outcome periods and dosage categories
outcome_periods <- c("1M", "3M", "6M", "12M", "18M")
dosage_categories <- c("max_dosage_1M_cat_grouped", "max_dosage_3M_cat_grouped", 
                       "max_dosage_6M_cat_grouped", "max_dosage_12M_cat_grouped",
                       "max_dosage_18M_cat_grouped")

# Initialize a list to store results
results_list <- list()

# Loop over each outcome period
for (i in seq_along(outcome_periods)) {
  outcome <- outcome_periods[i]
  dosage <- dosage_categories[i]

  # Fit the generalized linear model
  model_formula <- as.formula(
  paste0("outcome.", outcome, " ~ ", dosage, " + age_cat + race + PAT_GENDER_CD + payment + year")
)

# Fit the model
  result <- glm(model_formula, data = nochange, family = "binomial")
  
  # Extract coefficients and calculate odds ratios and confidence intervals
  coeff_df <- as.data.frame(summary(result)$coefficients)
  coeff_df <- coeff_df[-1,]  # Exclude the intercept
  
  or_df <- as.data.frame(
    cbind("Odds ratio" = exp(coef(result)), exp(confint.default(result)))
  )
  or_df <- or_df[-1,]  # Exclude the intercept
  
  # Combine the coefficient and odds ratio data frames
  out_coeff <- bind_cols(coeff_df, or_df)
  
  # Add a column indicating the outcome period
  out_coeff$Outcome_Period <- outcome
  
  # Store the result in the list
  results_list[[outcome]] = out_coeff
  
  # Optionally remove intermediate variables to save memory
  rm(result, coeff_df, or_df)
}

# Concatenate all data frames in the list
combined_results <- do.call(rbind, results_list)

# Write the combined results to a CSV file
combined_output_filename <- paste0(result.folder, "combined_multivariable_M18_nochangedosage_revision.csv")
fwrite(combined_results, combined_output_filename, row.names = TRUE)

# Optionally clean up memory
rm(combined_results, results_list, combined_output_filename, dosage, dosage_categories,
i, model_formula, outcome, out_coeff, outcome_periods)
```





Pairwise comparisons for each of the 10 dosage categories in multivariable logistic regression
```{r}
# Define the outcome periods and dosage categories
outcome_periods <- c("1M", "3M", "6M", "12M", "18M")
dosage_categories <- c("max_dosage_1M_cat_grouped", "max_dosage_3M_cat_grouped", 
                       "max_dosage_6M_cat_grouped", "max_dosage_12M_cat_grouped",
                       "max_dosage_18M_cat_grouped")

# Initialize a list to store results
pairwise_results_list <- list()

# Loop over each outcome period and compute pairwise comparisons
for (i in seq_along(outcome_periods)) {
  outcome <- outcome_periods[i]
  dosage <- dosage_categories[i]

  # Fit the generalized linear model
  model_formula <- as.formula(
    paste0("outcome.", outcome, " ~ ", dosage, " + age_cat + race + PAT_GENDER_CD + payment + year")
  )
  result <- glm(model_formula, data = nochange, family = "binomial")

  # Compute estimated marginal means for pairwise comparisons of dosage categories
  # Using reformulate to construct the correct variable name for emmeans
  emm <- emmeans(result, specs = reformulate(dosage))
  
  # Get pairwise comparisons and adjust for multiple comparisons (Tukey adjustment)
  pairwise_comp <- pairs(emm, adjust = "tukey")
  
  # Extract results and add a column indicating the outcome period
  pairwise_df <- as.data.frame(summary(pairwise_comp))
  
  # Calculate Odds Ratio and 95% Confidence Intervals
  pairwise_df$Odds_Ratio <- exp(pairwise_df$estimate)
  pairwise_df$Lower_CI <- exp(pairwise_df$estimate - 1.96 * pairwise_df$SE)
  pairwise_df$Upper_CI <- exp(pairwise_df$estimate + 1.96 * pairwise_df$SE)
  
  pairwise_df$Outcome_Period <- outcome
  
  # Store the result in the list
  pairwise_results_list[[outcome]] <- pairwise_df
  
  # Optionally remove intermediate variables to save memory
  rm(result, emm, pairwise_comp)
}

# Concatenate all data frames in the list
combined_pairwise_results <- do.call(rbind, pairwise_results_list)

# Write the pairwise comparison results to a CSV file
pairwise_output_filename <- paste0(result.folder, "pairwise_multivariable_M18_nochangedosage_revision.csv")
fwrite(combined_pairwise_results, pairwise_output_filename, row.names = TRUE)

# Optionally clean up memory
rm(combined_pairwise_results, pairwise_results_list, pairwise_output_filename,
   dosage, dosage_categories, i, model_formula, outcome, outcome_periods,pairwise_df)


```





Test interaction terms for paper revision
```{r}
# Fit the model
result.1M <- glm(outcome.1M ~ max_dosage_1M_cat_grouped*year + age_cat + race + PAT_GENDER_CD + payment, data = nochange, family = "binomial"); summary(result.1M)

result.3M <- glm(outcome.3M ~ max_dosage_3M_cat_grouped*year + age_cat + race + PAT_GENDER_CD + payment, data = nochange, family = "binomial"); summary(result.3M)

result.6M <- glm(outcome.6M ~ max_dosage_6M_cat_grouped*year + age_cat + race + PAT_GENDER_CD + payment, data = nochange, family = "binomial"); summary(result.6M)

result.12M <- glm(outcome.12M ~ max_dosage_12M_cat_grouped*year + age_cat + race + PAT_GENDER_CD + payment, data = nochange, family = "binomial"); summary(result.12M)

result.18M <- glm(outcome.18M ~ max_dosage_18M_cat_grouped*year + age_cat + race + PAT_GENDER_CD + payment, data = nochange, family = "binomial"); summary(result.18M)

sink(file =paste0(result.folder,"Interaction output.txt"))
summary(result.1M)
summary(result.3M)
summary(result.6M)
summary(result.12M)
summary(result.18M)
sink(file = NULL)
```





Stratify the multivariable logistic regression by year
```{r}
# Define the outcome periods and dosage categories
outcome_periods <- c("1M", "3M", "6M", "12M", "18M")
dosage_categories <- c("max_dosage_1M_cat_grouped", "max_dosage_3M_cat_grouped", 
                       "max_dosage_6M_cat_grouped", "max_dosage_12M_cat_grouped",
                       "max_dosage_18M_cat_grouped")

# Initialize a list to store results
results_list <- list()

# Loop over each year
for (Y in 2014:2019) {
  
  # Filter the dataset for the current year
  ifelse(Y<2019, year_data <- nochange[nochange$year== Y, ],year_data <- nochange[nochange$year>=Y, ])
 
  
  # Loop over each outcome period
  for (i in seq_along(outcome_periods)) {
    outcome <- outcome_periods[i]
    dosage <- dosage_categories[i]
    
    # Fit the generalized linear model
    model_formula <- as.formula(
      paste0("outcome.", outcome, " ~ ", dosage, " + age_cat + race + PAT_GENDER_CD + payment")
    )
    
    # Fit the model
    result <- glm(model_formula, data = year_data, family = "binomial")
    
    # Extract coefficients and calculate odds ratios and confidence intervals
    coeff_df <- as.data.frame(summary(result)$coefficients)
    coeff_df <- coeff_df[-1, ]  # Exclude the intercept
    
    or_df <- as.data.frame(
      cbind("Odds ratio" = exp(coef(result)), exp(confint.default(result)))
    )
    or_df <- or_df[-1, ]  # Exclude the intercept
    
    # Combine the coefficient and odds ratio data frames
    out_coeff <- bind_cols(coeff_df, or_df)
    
    # Add columns indicating the outcome period and year
    out_coeff$Outcome_Period <- outcome
    out_coeff$Year <- Y
    
    # Store the result in the list
    results_list[[paste0(outcome, "_", Y)]] <- out_coeff
    
    # Optionally remove intermediate variables to save memory
    rm(result, coeff_df, or_df)
  }
}

# Concatenate all data frames in the list
combined_results <- do.call(rbind, results_list)

# Write the combined results to a CSV file
combined_output_filename <- paste0(result.folder, "combined_multivariable_M18_nochangedosage_byyear_revision2.csv")
fwrite(combined_results, combined_output_filename, row.names = TRUE)

# Optionally clean up memory
rm(combined_results, results_list, combined_output_filename, dosage, dosage_categories,
   i, model_formula, outcome, out_coeff, outcome_periods, year_data)

```





```{r}
year_data <- nochange[nochange$year == 2019, ]
result <- glm(outcome.12M ~ max_dosage_12M_cat_grouped + age_cat + race + PAT_GENDER_CD + payment, 
                               data = year_data, family = "binomial")
summary(result)
```



